### 2016-12-09 greedy_hamiltonian_revised/chapters/4.sympmor.tex
Jan's comments included

\section{Symplectic Model Reduction} \label{chap:SyMo:1}
We now propose how to modify reduced order modeling to ensure that they preserve the symplectic structure of Hamiltonian system.

Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $(P,\Omega)$. Suppose that the solution manifold $\mathcal M_H$ is well approximated, by a low dimensional symplectic subspace $(Q,\Omega)$ of dimension $2k$ $(k\ll n)$. We can construct a symplectic basis $A$ for $Q$ and approximate the solution to (\ref{eq:Hasy:7}) as
\begin{equation} \label{eq:SyMo:1}
	\mathbf z \approx A\mathbf y.
\end{equation}
Substituting this into (\ref{eq:Hasy:7}) we obtain
\begin{equation} \label{eq:SyMo:2}
	A\mathbf y = \mathbb{J}_{2n} \nabla_{\mathbf z} H(A \mathbf y). 
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have
\begin{equation} \label{eq:SyMo:3}
	\mathbf y = A^+ \mathbb J_{2n} (A^+)^T \nabla_{\mathbf y} H(A\mathbf y).
\end{equation}
Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e. $A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H:\mathbb R^{2k} \to \mathbb R$ as $\tilde H (y) = H(Ay)$ we obtain the reduced system
\begin{equation} \label{eq:SyMo:4}
\left\{
\begin{aligned}
	 \frac{d}{dt} \mathbf y &= \mathbb J_{2k} \nabla_{\mathbf y} \tilde H(\mathbf y), \\
	 \mathbf y_0 &= A^+ \mathbf z_0.
\end{aligned}
\right.
\end{equation}
The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and therefore does not guarantee conservation of the symplectic structure. On the other hand we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e. the symplectic structure will be conserved along integral curves of (\ref{eq:SyMo:4}). Note that the original and the reduced systems are derived from different Hamiltonians. In the next proposition we show that the error in the Hamiltonian is constant in time.

@@

\begin{definition} \label{definition:SyMo:1}
Consider a dynamical system of the form $\dot{\mathbf z} = \mathbf f(\mathbf z)$ and suppose that $\mathbf z_e$ is an equilibrium point for the system so that $\mathbf f(\mathbf z_e) = 0$. $\mathbf z_e$ is called nonlinear stable or Lyapunov stable if for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\phi_t$, if $\| \phi_0 - \mathbf z_e \|_2 \leq \delta$, then for all $t\geq0$, we have $\| \phi_t - \mathbf z_e \|_2 < \epsilon$, where $\| \cdot \|_2$ is the Euclidean norm.
\end{definition}	
The following proposition, also known as Dirichlet's theorem \cite{bhatia2002stability}, states the sufficient condition for an equilibrium point to be Lyapunov stable. We refer the reader to \cite{bhatia2002stability} for the proof.
\begin{proposition} \label{proposition:SyMo:1}
An equilibrium point $\mathbf z_e$ is Lyapunov stable if there exist a scalar function $W : \mathbb R^{n} \to  \mathbb R$ such that $\nabla W(\mathbf z_e) = 0$, and $\nabla^2 W(\mathbf z_e)$ is positive definite and that for any trajectory $\phi_t$ defined in the neighborhood of $\mathbf z_e$, we have $\frac{d}{dt} W(\phi_t) \leq 0$. Here $\nabla^2W$ is the Hessian matrix of $W$.
\end{proposition}
The scalar function $W$ is referred to as the \emph{Lyapunov function}. In the context of the Hamiltonian systems, a suitable candidate for the Lyapunov function is the Hamiltonian function $H$. The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Liyapunov stable \cite{abraham1978foundations}. 
\begin{theorem} \label{theorem:SyMo:1}
Consider a Hamiltonian system of the form (\ref{eq:Hasy:7}) together with the reduced system (\ref{eq:SyMo:4}). Suppose $\mathbf z_e$ is an equilibrium point for (\ref{eq:Hasy:7}) and that $\mathbf y_e = A^+\mathbf z_e$. If $H$ (or $-H$) is a Lyapunov function satisfying proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable equilibrium points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}) respectively. 
\end{theorem}
\begin{proof}
	It is the direct consequence of proposition \ref{proposition:SyMo:1} that $\mathbf z_e$ is a local minimum or maximum of (\ref{eq:Hasy:7}) and also a Lyapunov stable point. It can be easily checked that if $\mathbf z_e$ is a local minimum of $H$ then $\mathbf y_e$ is a local minimum for $\tilde H$ and an equilibrium point for (\ref{eq:SyMo:4}). Also from the chain rule we have
\[
	\nabla^2_{\mathbf y} \tilde H = A^T \nabla^2_{\mathbf z} H A.
\]
So for any $\xi\in \mathbb R^{2k}$
\[
	\xi^T \nabla^2_{\mathbf y} \tilde H \xi = (A\xi)^T \nabla^2_{\mathbf z} H (A\xi) \geq 0,
\]
Where the last inequality is due to the positive definiteness of $H$. Therefore $\tilde H$ is also positive definite. By proposition \ref{proposition:SyMo:1} we conclude that $\mathbf y_e$ is a Lyapunov stable point.
\end{proof}
}

@@

Similar to Section \ref{chap:MoOr.PrOr:1} we gather snapshots $\mathbf z_i = [q_i^T , p_i^T]^T$ in the snapshot matrix $S$. Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided. {\edit The Proper Symplectic Decomposition} requires that the error of the symplectic projection onto the symplectic subspace to be minimum. Analogous to the minimization (\ref{eq:MoOr:6}), the PSD symplectic basis of size $2k$ is the solution of the optimization problem

\begin{equation} \label{eq:SyMo:8}
\begin{aligned}
& \underset{V\in \mathbb R^{2n\times 2k}}{\text{minimize}}
& & \| S - AA^+S\|_F \\
& \text{subject to}
& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}
\end{aligned}
\end{equation}
Compared to POD, in (\ref{eq:SyMo:8}) the orthogonal projection is replaced with a symplectic projection $AA^+$. At first, the minimization looks similar to the one obtained by POD. {\edit However, it is well known that symplectic bases are not generally orthogonal, and therefore not norm bounded, this means numerical errors can become dominant in symplectic projection \cite{Karow:2006cf}. This makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).}

As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, direct solution usually expensive. A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis.

{\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices is proposed by \cite{Xu:2003kx}. However, computation of a possibly large snapshot matrix and then using the mentioned methods to compute its eigen-spaces, is usually computationally demanding. Also, these methods generally do not guarantee the construction of a well-conditioned symplectic basis.

The greedy approach presented in section \ref{Chap:Symo.PrSy:3}, is an iterative method for construction of a symplectic basis. It avoids evaluation of the full snapshot matrix so it saves substantial computational cost in the offline stage of the symplectic model reduction. Also, by construction, it yields a orthosymplectic basis and therefore a well-conditioned basis.

In section \ref{chap:SyMo.PrSy:2} we outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, while assuming a specific structure for $A$. Finally, in section \ref{Chap:Symo.PrSy:3} we introduce a greedy approach for symplectic basis generation.}

@@

\paragraph{\bf Complex SVD} We now consider another form of symplectic basis $A$. Suppose that $A$ takes the form \cite{Peng:2014di}
\begin{equation} \label{eq:SyMo:11}
	A = 
	\begin{pmatrix}
		\Phi & -\Psi \\
		\Psi & \Phi
	\end{pmatrix},
\end{equation}
such that $\Phi$ and $\Psi$ are real matrices of the size $n\times k$ satisfying conditions
\begin{equation} \label{eq:SyMo:12}
\Phi^T \Phi + \Psi^T \Psi = I_k,\quad \Phi^T \Psi = \Psi^T \Phi.
\end{equation}
It can be checked that $A$ forms a symplectic matrix. To construct $A$ we first define the complex snapshot matrix
\begin{equation} \label{eq:SyMo:13}
	S_{\text{complex}} = [ q_1 + i p_1, \dots , q_N + i p_N ].
\end{equation}
The SVD decomposition of $S_{\text{complex}}$ can be written as $S_{\text{complex}} = U \Sigma V^H$ with the superscript $H$ indicating the conjugate transpose of a complex valued matrix. Each left singular vector of $S_{\text{complex}}$ takes the form $u_m = r_m + i s_m$. We define
\begin{equation} \label{eq:SyMo:14}
	 \Phi = [r_1,\dots, r_k], \quad \Psi = [s_1,\dots, s_k].
\end{equation}
Then we can easily check that conditions (\ref{eq:SyMo:12}) is satisfied since $U$ is unitary. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:11}) the complex SVD minimizes the projection error.

\subsubsection{The Greedy Approach to Symplectic Basis Generation} \label{Chap:Symo.PrSy:3} greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This results in more compact bases and/or more accurate reduced systems. In addition, for parametric problems, the greedy approach only requires one numerical solution to be computed per iteration which saves substantial computational cost in the offline stage.

{\edit The orthonormalization step is an essential step in most greedy approaches for basis generation in the context of model reduction \cite{Anonymous:2016wl,Quarteroni:2016wi}. However common orthonormalization processes, e.g. the QR method based on the classical Gram-Schmidt method, destroy the symplectic structure of the original system \cite{BunseGerstner:1986dg}. Here we use a variation of the QR method known as the the SR \cite{Salam2014} method which is based on the symplectic Gram-Schmidt method and yields a symplectic basis. 
}

{\edit
As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}). Further, theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}. To briefly describe the SR method, suppose that an orthosymplectic basis
\begin{equation} \label{eq:SyMo:14.1}
	A_{2k}=\{ e_1 , \dots , e_k \} \cup \{ \mathbb J_{2n}^T e_1 , \dots , \mathbb J_{2n}^T e_k \},
\end{equation}
and a vector $z\not \in \text{span}(A_{2k})$ is provided. We would like to symplectically orthogonalize ($\mathbb J_{2n}$-orthogonalize) $z$ with respect to $A_{2k}$. We seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation} \label{eq:SyMo:14.2}
	\Omega\left(z + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i \mathbb J_{2n}^Te_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i \mathbb J_{2n}^Te_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is as 
\begin{equation} \label{eq:SyMo:14.3}
	\alpha_i = - \Omega(z,\mathbb J_{2n}^Te_i), \quad \beta_i = \Omega(z,e_i),
\end{equation}
for $i=1,\dots,k$. Now we define the modified vectors as
\begin{equation} \label{eq:SyMo:14.4}
	\tilde z = z - \sum_{i=1}^k \Omega(z,\mathbb J_{2n}^Te_i) e_i + \sum_{i=1}^k \Omega(z,e_i) \mathbb J_{2n}^Te_i.
\end{equation}
If we introduce $e_{k+1} = \tilde z / \| \tilde z \|_2$, it is easily checked that $e_{k+1}$ is also orthogonal to $A_{2k}$ with respect to the classical inner product. Therefore the set span$(\{e_1,\dots,e_{k+1}\})$ forms a Lagrangian subspace and according to theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^T e_{k+1} \}$ forms an orthosymplectic basis.

Note that the $SR$ method can be replaced with a backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.
}

The key element of the greedy algorithm is the availability of an error function which evaluates the error associated with the model reduction \cite{Anonymous:2016wl}. In the framework of symplectic model reduction, one possible candidate is the error in the Hamiltonian (\ref{eq:SyMo:5}). Correctly approximating symplectic systems relies on preservation of the Hamiltonian, hence the error in the Hamiltonian appears to be a natural choice. Moreover, since the error in the Hamiltonian depends on the initial condition and the reduced symplectic basis, evaluation of error does not require time integration of the full system.

Suppose that a $2k$ dimensional {\edit orthosymplectic basis (\ref{eq:SyMo:14.1})} is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors. Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to find the value that maximises the error in the Hamiltonian
\begin{equation} \label{eq:SyMo:14.5}
	\omega_{k+1} := \underset{\omega\in \Gamma}{\text{argmax }}\Delta H(\omega).
\end{equation}
The goal is to approximate the Hamiltonian function as well as possible.

We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots 
\begin{equation}
	S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}.
\end{equation} 
The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
{\edit 
\begin{equation} \label{eq:SyMo:14.6}
	z := \underset{s\in S}{\text{argmax }} \| s - A_{2k}{A_{2k}}^+s \|.
\end{equation}
}
Finally, we update the basis as
{\edit
\begin{equation} \label{eq:SyMo:14.7}
	e_{k+1} = \tilde z, \quad A_{2k+1} = A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^Te_{k+1} \},
\end{equation}
}
where $\tilde z$ is the vector obtained from applying the symplectic Gram-Schmidt process to $z$.

Since the maximization over the entire parameter space $\Gamma$ is impossible, we discretize the parameter set into a grid with $N$ points: $\Gamma_N = \{ \omega_1,\dots,\omega_N\}$. However, since the selection of parameters only require the evaluation of the error in the Hamiltonian and not time integration of the original system, then $\Gamma_N$ can be chosen very rich. Other sampling techniques can be found in \cite{Quarteroni:2016wi}.

We summarized the greedy algorithm for generation of a symplectic basis in Algorithm \ref{alg:SyMo:3}.

>>>>

\section{Symplectic Model Reduction} \label{chap:SyMo:1}
We now {\edit discuss} how to modify reduced order modeling to ensure that {\edit the resulting scheme preserves} the symplectic structure of the Hamiltonian system.

Consider a Hamiltonian system (\ref{eq:Hasy:7}) on a $2n$-dimensional symplectic vector space $\edit (V,\Omega)$. Suppose that the solution manifold $\mathcal M_H$ is well approximated by a low dimensional symplectic subspace $\edit (W,\Omega)$ of dimension $2k$ $(k\ll n)$. We can {\edit then} construct a symplectic basis $A$ for $\edit W$ and approximate the solution to (\ref{eq:Hasy:7}) as
\begin{equation} \label{eq:SyMo:1}
	\mathbf z \approx A\mathbf y.
\end{equation}
Substituting this into (\ref{eq:Hasy:7}) we obtain
\begin{equation} \label{eq:SyMo:2}
	A\mathbf y = \mathbb{J}_{2n} \nabla_{\mathbf z} H(A \mathbf y). 
\end{equation}
Multiplying both sides with the symplectic inverse of $A$ and using the chain rule we have
\begin{equation} \label{eq:SyMo:3}
	\mathbf y = A^+ \mathbb J_{2n} (A^+)^T \nabla_{\mathbf y} H(A\mathbf y).
\end{equation}
Since $A$ is a symplectic basis, Lemma \ref{lemma:Hasy:1} ensures that $(A^+)^T$ is a symplectic matrix i.e., $A^+ \mathbb J_{2n} (A^+)^T = \mathbb{J}_{2k}$. By defining the reduced Hamiltonian $\tilde H:\mathbb R^{2k} \to \mathbb R$ as $\tilde H (y) = H(Ay)$ we obtain the reduced system
\begin{equation} \label{eq:SyMo:4}
\left\{
\begin{aligned}
	 \frac{d}{dt} \mathbf y &= \mathbb J_{2k} \nabla_{\mathbf y} \tilde H(\mathbf y), \\
	 \mathbf y_0 &= A^+ \mathbf z_0.
\end{aligned}
\right.
\end{equation}
The system obtained from the Petrov-Galerkin projection in (\ref{eq:MoOr:5}) is not a Hamiltonian system and does not guarantee conservation of the symplectic structure. On the other hand, we observe that the reduced system in (\ref{eq:SyMo:4}) is of the form (\ref{eq:Hasy:7}) and, hence, is a Hamiltonian system, i.e. the symplectic structure will be conserved along integral curves of (\ref{eq:SyMo:4}). Note that the original and the reduced systems are {\edit endowed with} different Hamiltonians. In the next proposition we show that the error in the Hamiltonian is constant in time.

@@

\begin{definition} \label{definition:SyMo:1} \cite{bhatia2002stability}
Consider a dynamical system of the form $\dot{\mathbf z} = \mathbf f(\mathbf z)$ and suppose that $\mathbf z_e$ is an equilibrium point for the system so that $\mathbf f(\mathbf z_e) = 0$. $\mathbf z_e$ is called nonlinearly stable or Lyapunov stable if, for any $\epsilon > 0$, we can find $\delta > 0$ such that for any trajectory $\phi_t$, if $\| \phi_0 - \mathbf z_e \|_2 \leq \delta$, then for all $0 \leq t < \infty$, we have $\| \phi_t - \mathbf z_e \|_2 < \epsilon$, where $\| \cdot \|_2$ is the Euclidean norm.
\end{definition}	
The following proposition, also known as Dirichlet's theorem \cite{bhatia2002stability}, states the sufficient condition for an equilibrium point to be Lyapunov stable. We refer the reader to \cite{bhatia2002stability} for the proof.
\begin{proposition} \label{proposition:SyMo:1} \cite{bhatia2002stability}
An equilibrium point $\mathbf z_e$ is Lyapunov stable if there exists a scalar function $W : \mathbb R^{n} \to  \mathbb R$ such that $\nabla W(\mathbf z_e) = 0$, $\nabla^2 W(\mathbf z_e)$ is positive definite, and that for any trajectory $\phi_t$ defined in the neighborhood of $\mathbf z_e$, we have $\frac{d}{dt} W(\phi_t) \leq 0$. Here $\nabla^2W$ is the Hessian matrix of $W$.
\end{proposition}
The scalar function $W$ is referred to as the \emph{Lyapunov function}. In the context of the Hamiltonian systems, a suitable candidate for the Lyapunov function is the Hamiltonian function $H$. The following theorem shows that when $H$ (or $-H$) is a Lyapunov function, then the equilibrium points of the original and the reduced system are Lyapunov stable \cite{abraham1978foundations}. 
\begin{theorem} \label{theorem:SyMo:1}
Consider a Hamiltonian system of the form (\ref{eq:Hasy:7}) together with the reduced system (\ref{eq:SyMo:4}). Suppose $\mathbf z_e$ is an equilibrium point for (\ref{eq:Hasy:7}) and that $\mathbf y_e = A^+\mathbf z_e$. If $H$ (or $-H$) is a Lyapunov function satisfying Proposition \ref{proposition:SyMo:1}, then $\mathbf z_e$ and $\mathbf y_e$ are Lyapunov stable equilibrium points for (\ref{eq:Hasy:7}) and (\ref{eq:SyMo:4}), respectively. 
\end{theorem}
\begin{proof}
	It is a direct consequence of Proposition \ref{proposition:SyMo:1} that $\mathbf z_e$ is a local minimum or maximum of (\ref{eq:Hasy:7}) and also a Lyapunov stable point. It can be easily checked that if $\mathbf z_e$ is a local minimum of $H$ then $\mathbf y_e$ is a local minimum for $\tilde H$ and an equilibrium point for (\ref{eq:SyMo:4}). Also from the chain rule we have
\[
	\nabla^2_{\mathbf y} \tilde H = A^T \nabla^2_{\mathbf z} H A.
\]
So for any $\xi\in \mathbb R^{2k}$
\[
	\xi^T \nabla^2_{\mathbf y} \tilde H \xi = (A\xi)^T \nabla^2_{\mathbf z} H (A\xi) \geq 0.
\]
Here the last inequality is due to the positive definiteness of $H$. Therefore $\tilde H$ is also positive definite. By Proposition \ref{proposition:SyMo:1} we conclude that $\mathbf y_e$ is a Lyapunov stable point.
\end{proof}
}

@@

Similar to Section \ref{chap:MoOr.PrOr:1} we gather snapshots $\mathbf z_i = [q_i^T , p_i^T]^T$ in the snapshot matrix $S$. Suppose that a symplectic basis $A$ of size $2n\times2k$ and its symplectic inverse $A^+$ is provided. {\edit The Proper Symplectic Decomposition} requires that the error of the symplectic projection onto the symplectic subspace {\edit be minimized}. Hence, the PSD symplectic basis of size $2k$ is the solution to the optimization problem

\begin{equation} \label{eq:SyMo:8}
\begin{aligned}
& \underset{V\in \mathbb R^{2n\times 2k}}{\text{minimize}}
& & \| S - AA^+S\|_F \\
& \text{subject to}
& & A^T \mathbb{J}_{2n}A = \mathbb{J}_{2k}
\end{aligned}
\end{equation}
Compared to POD, in (\ref{eq:SyMo:8}) the orthogonal projection is replaced with a symplectic projection $AA^+$. At first, the minimization looks similar to the one obtained by POD. {\edit However, it is well known that symplectic bases are not generally orthogonal, and therefore not norm bounded. This means that numerical errors may become dominant in the symplectic projection \cite{Karow:2006cf} which makes the minimization (\ref{eq:SyMo:8}) a harder problem than (\ref{eq:MoOr:6}).}

As the optimization problem (\ref{eq:SyMo:8}) is nonlinear, the direct solution is usually expensive. A simplified version of the optimization (\ref{eq:SyMo:8}) can be found in \cite{Peng:2014di}, but there is no guarantee that the method provides a near optimal basis.

{\edit Finding eigen-spaces of Hamiltonian and symplectic matrices is studied in the context of optimal control problems \cite{Benner:2000ww,Benner:1997ef,Watkins:2004kz,BunseGerstner:1986dg} and model reduction of Riccati equations \cite{Benner:1997ef}, where also an SVD-like decomposition for Hamiltonian and symplectic matrices has been proposed \cite{Xu:2003kx}. However, the computation of a large snapshot matrix and use of the mentioned methods to compute its eigen-spaces, is usually computationally demanding. Also, these methods generally do not guarantee the construction of a well-conditioned symplectic basis.

The greedy approach presented in Section \ref{Chap:Symo.PrSy:3} is an iterative method for construction of a symplectic basis. It avoids the evaluation of the full snapshot matrix, hence substantially reduces the computational cost in the offline stage of the symplectic model reduction. Also, by construction, it yields a orthosymplectic basis and therefore a well-conditioned basis.

In Section \ref{chap:SyMo.PrSy:2} we briefly outline non-direct methods for finding solutions to (\ref{eq:SyMo:8}), proposed by \cite{Peng:2014di}, and assuming a specific structure for $A$. In Section \ref{Chap:Symo.PrSy:3} we introduce a greedy approach for the symplectic basis generation.}

@@

\paragraph{\bf Complex SVD} Suppose instead that $A$ takes the form \cite{Peng:2014di}
\begin{equation} \label{eq:SyMo:11}
	A = 
	\begin{pmatrix}
		\Phi & -\Psi \\
		\Psi & \Phi
	\end{pmatrix},
\end{equation}
while $\Phi$ and $\Psi$ are real matrices of size $n\times k$ satisfying conditions
\begin{equation} \label{eq:SyMo:12}
\Phi^T \Phi + \Psi^T \Psi = I_k,\quad \Phi^T \Psi = \Psi^T \Phi.
\end{equation}
It can be checked that $A$ forms a symplectic matrix. To construct $A$ we first define the complex snapshot matrix
\begin{equation} \label{eq:SyMo:13}
	S_{\text{complex}} = [ q_1 + i p_1, \dots , q_N + i p_N ].
\end{equation}
Each left singular vector of $S_{\text{complex}}$ now takes the form $u_m = r_m + i s_m$. We define
\begin{equation} \label{eq:SyMo:14}
	 \Phi = [r_1,\dots, r_k], \quad \Psi = [s_1,\dots, s_k].
\end{equation}
One can easily check that (\ref{eq:SyMo:12}) is satisfied {\edit since the matrix of singular vectors is unitary}. It is shown in \cite{Peng:2014di} that among all symplectic bases of the form (\ref{eq:SyMo:11}) the complex SVD minimizes the projection error.

\subsubsection{The Greedy Approach to Symplectic Basis Generation} \label{Chap:Symo.PrSy:3} Greedy generation of the reduced basis is an iterative procedure which, in each iteration, adds the two best possible basis vectors to the symplectic basis to enhance overall accuracy. In contrast to the cotangent lift and the complex SVD methods, the greedy approach does not require the symplectic basis to have a specific structure. This typically results in a more compact basis and/or more accurate reduced systems. For parametric problems, the greedy approach only requires one numerical solution to be computed per iteration hence saving substantial computational cost in the offline stage.

{\edit The orthonormalization step is an essential step in most greedy approaches for basis generation in the context of model reduction \cite{Anonymous:2016wl,Quarteroni:2016wi}. However common orthonormalization processes, e.g. the QR method, destroy the symplectic structure of the original system \cite{BunseGerstner:1986dg}. Here we use a variation of the QR method known as the SR \cite{Salam2014} method which is based on the symplectic Gram-Schmidt method and yields a symplectic basis. 
}

{\edit
As discussed in Section \ref{chap:Hasy:1}, any finite dimensional symplectic linear vector space has a symplectic basis that satisfies conditions (\ref{eq:Hasy:4}). Further, Theorem \ref{theorem:Hasy:1.6} provides an iterative process for constructing an orthosymplectic basis \cite{Matsuo:2014wl,Salam2014}. To briefly describe the SR method, suppose that an orthosymplectic basis
\begin{equation} \label{eq:SyMo:14.1}
	A_{2k}=\{ e_1 , \dots , e_k \} \cup \{ \mathbb J_{2n}^T e_1 , \dots , \mathbb J_{2n}^T e_k \},
\end{equation}
and a vector $z\not \in \text{span}(A_{2k})$ is provided. We aim to symplectically orthogonalize ($\mathbb J_{2n}$-orthogonalize) $z$ with respect to $A_{2k}$ and seek $\alpha_1,\dots,\alpha_k,\beta_1,\dots,\beta_k \in \mathbb R$ such that
\begin{equation} \label{eq:SyMo:14.2}
	\Omega\left(z + \sum_{i=1}^k \alpha_i e_i + \sum_{i=1}^k \beta_i \mathbb J_{2n}^Te_i , \sum_{i=1}^k \bar{\alpha}_i e_i + \sum_{i=1}^k \bar{\beta}_i \mathbb J_{2n}^Te_i \right) = 0,
\end{equation}
for all possible $\bar{\alpha}_1,\dots,\bar{\alpha}_k,\bar{\beta}_1,\dots,\bar{\beta}_k \in \mathbb R$. It is easily seen that the unique solution is 
\begin{equation} \label{eq:SyMo:14.3}
	\alpha_i = - \Omega(z,\mathbb J_{2n}^Te_i), \quad \beta_i = \Omega(z,e_i),
\end{equation}
for $i=1,\dots,k$. Now define the modified vectors as
\begin{equation} \label{eq:SyMo:14.4}
	\tilde z = z - \sum_{i=1}^k \Omega(z,\mathbb J_{2n}^Te_i) e_i + \sum_{i=1}^k \Omega(z,e_i) \mathbb J_{2n}^Te_i.
\end{equation}
If we introduce $e_{k+1} = \tilde z / \| \tilde z \|_2$, it is easily checked that $e_{k+1}$ is also orthogonal to $A_{2k}$ with respect to the classical inner product. Therefore span$\{e_1,\dots,e_{k+1}\}$ forms a Lagrangian subspace and according to Theorem \ref{theorem:Hasy:1.6} the basis $A_{2k+2}= A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^T e_{k+1} \}$ forms an orthosymplectic basis.

Note that the $SR$ method can be replaced with backward stable routines such as the isotropic Arnoldi or the isotropic Lanczos methods \cite{Mehrmann:2000dv}.
}

The key element of the greedy algorithm is the availability of an error function which evaluates the error associated with the model reduction \cite{Anonymous:2016wl}. In the framework of symplectic model reduction, one possible candidate is the error in the Hamiltonian (\ref{eq:SyMo:5}). Correctly approximating symplectic systems relies on preservation of the Hamiltonian, hence the error in the Hamiltonian {\edit arises as a} a natural choice. Moreover, since the error in the Hamiltonian depends on the initial condition and the reduced symplectic basis, evaluation of the error does not require the time integration of the full system.

Suppose that a $2k$-dimensional {\edit orthosymplectic basis (\ref{eq:SyMo:14.1})} is generated at the $k$-th step of the greedy method and we seek to enrich it by two additional vectors. Using the error in the Hamiltonian (\ref{eq:SyMo:7}) we search the parameter space to identify the value that maximizes the error in the Hamiltonian
\begin{equation} \label{eq:SyMo:14.5}
	\omega_{k+1} := \underset{\omega\in \Gamma}{\text{argmax }}\Delta H(\omega).
\end{equation}
The goal is to approximate the Hamiltonian function as well as possible.

We then propagate (\ref{eq:Hasy:7}) in time to produce trajectory snapshots 
\begin{equation}
	S=\{ \mathbf z(t_i,\omega_{k+1}) | i = 1,\dots,M \}.
\end{equation} 
The next basis vector is the snapshot that maximises the projection error (\ref{eq:SyMo:8})
{\edit 
\begin{equation} \label{eq:SyMo:14.6}
	z := \underset{s\in S}{\text{argmax }} \| s - A_{2k}{A_{2k}}^+s \|.
\end{equation}
}
Finally, we update the basis as
{\edit
\begin{equation} \label{eq:SyMo:14.7}
	e_{k+1} = \tilde z, \quad A_{2k+1} = A_{2k}\cup \{ e_{k+1} , \mathbb J_{2n}^Te_{k+1} \},
\end{equation}
}
where $\tilde z$ is the vector obtained {\edit after} applying the symplectic Gram-Schmidt process to $z$.

Since the maximization over the entire parameter space $\Gamma$ is impossible, we discretize the parameter set into a grid with $N$ points: $\Gamma_N = \{ \omega_1,\dots,\omega_N\}$. However, since the selection of parameters only require the evaluation of the error in the Hamiltonian and not time integration of the original system, then $\Gamma_N$ can be chosen {\edit to be} very rich.

We summarize the greedy algorithm for the generation of a symplectic basis in Algorithm \ref{alg:SyMo:3}.

\begin{algorithm} 
\caption{The greedy algorithm for generation of a symplectic basis} \label{alg:SyMo:3}
{\bf Input:} Tolerated loss in the Hamiltonian $\delta$, parameter set $\Gamma_N = \{\omega_1,\dots,\omega_N\}$, initial condition $\mathbf z_0(\omega)$
\begin{enumerate}
\item $\omega^* \leftarrow \omega_1$
\item $e_1 \leftarrow \mathbf z_0(\omega^*)$
\item $A \leftarrow [e_1,-\mathbb J^T_{2n}e_1]$
\item $k \leftarrow 1$
\item \textbf{while} $\Delta H(\omega) > \delta$ for all $\omega \in \Omega_N$
\item \hspace{0.5cm} $w^* \leftarrow$ $\underset{\omega\in \Omega_N}{\text{argmax }}\Delta H(\omega)$
\item \hspace{0.5cm} Compute trajectory snapshots $S=\{ \mathbf z(t_i,\omega^*) | i = 1,\dots,M \}$
\item \hspace{0.5cm} $\mathbf z^* \leftarrow$ $\underset{s\in S}{\text{argmax }} \| s - AA^+s \|$
\item \hspace{0.5cm} Apply symplectic Gram-Schmidt on $\mathbf z^*$
\item \hspace{0.5cm} $e_{k+1} \leftarrow \mathbf z^*/ \| \mathbf  z^*\|$
\item \hspace{0.5cm} $A \leftarrow [e_1,\dots ,e_{k+1} , \mathbb J^T_{2n}e_1,\dots,\mathbb J^T_{2n}e_{k+1}]$
\item \hspace{0.5cm} $k \leftarrow k+1$
\item \textbf{end while}
\end{enumerate}
\vspace{0.5cm}
{\bf Output:} Symplectic basis $A$.
\end{algorithm}
